Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-08-11 paper/paper.tex
Double checking paper results with other satellites

The top panel of Figure~\ref{ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot} along with that for 1--day and 1--hour medians.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used combined measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites is 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29, respectively.

@@

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 75 events from 1989-1991.  The minimum $D_{st}$ median is $-75$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $4.5$ amu/cm$^3$ instead of the increase of ${\sim} 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  Error lines for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values. For the non-$\rho_{eq}$\ parameters, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of $\rho_{eq}$\ values as all available measurements were used instead of restricting to only values where a $\rho_{eq}$ value also existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents}, it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.

To determine if the magnitude of the density medians in the top panel of Figure~\ref{DailyAveragedDstEvents} on epoch days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap sample of the medians for each epoch day was created by sampling the values used to compute the median on each epoch day with replacement.  The observed difference between the medians is shown in Table~\ref{BootstrapDifferenceTable} along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.

@@

A similar hypothesis test run for the bottom panel of Figure~\ref{DailyAveragedDstEvents} yields a maximum confidence level of $\sim$85\%.

@@

\begin{table}
\small
\centering
\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of means of $\rho_{eq}$ shown in the top panel of Figure~\ref{DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

We have also considered the hourly averages from the events shown in the bottom panel of Figure~\ref{DailyAveragedDstEvents}, which shown in the top panel of Figure~\ref{HourlyAveragedDstEvents}.  In the time period of elevated mass density within 6 hours of the start of the event, there are only $\sim$4 samples per hourly interval and the peak value of $\sim$70 amu/cm$^3$ is associated with a single observation.

To increase the number of events, we have removed the constraint that the events must occur in the MLT range from 06:00-12:00 and the constraint that $D_{st}$ remained below the threshold value for 12 hours or longer.  The result is shown in the bottom panel of Figure~\ref{HourlyAveragedDstEvents}. The two spikes (one upwards after onset, one downwards before onset) were tested for significant deviation from two baselines, one 24-12 hours before onset, one 16-48 hours after onset. The probability of the left spike coming from the same distribution as the left baseline is 1.06\%, and 8.2\% from the right baseline, whereas the probability of the right spike coming from the left baseline is 1.89\% and 0.17\% from the right baseline.  This indicates that there is a statistically significant peak in $\rho_{eq}$ within 6 hours of onset when hourly averages are considered.  This result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

To determine if there is an IMF $B_{z}$ dependence, the epoch curves in the bottom panel of Figure~\ref{HourlyAveragedDstEvents} for $\rho_{eq}$ and $B_{z}$ were separated into two parts depending whether $B_z$ was above or below its average four hours before and including the onset hour and four hours after and including the hour of onset; the result is shown in Figure~\ref{fig:BzBinnedDstEvents}. Any statistically significant difference at a 95\% confidence level between the two parts is indicated by a green dot on the horizontal axis using the same bootstrap method described previously.  (At this confidence level, false positives are expected to randomly occur 5\% of the time, or 3.6 times for 72 values.)  These two plots indicate that there is not a statistically significant dependence on IMF $B_{z}$ around the time of the onset of a $D_{st}$ event.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho_{eq}$.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\rho_{eq}$ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, missing values were replaced with linearly interpolated values.  These events are associated with positive $B_z$ and low values of $D_{st}$.  Approximately 12 hours after onset, $\rho_{eq}$ slowly decays over 40 hours to near pre--onset levels.

Figure~\ref{HourlyAveragedRhoEvents} can be compared with Figure~9 of \cite{Denton2016}, which shows daily averages of $\rho_{eq}$ from this dataset after the onset of quiet $K_p$ intervals.  The growth shown here is approximately linear and lasts $\sim$ 12--18 hours whereas the 10 quiet $K_p$ events considered by \cite{Denton2016} showed growth that lasted for at least 48~hours.

Figure~\ref{BzBinned} shows the the events of Figure~\ref{HourlyAveragedRhoEvents} when separated by $B_z$ in the same manner as Figure~\ref{fig:BzBinnedDstEvents}.  The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to a larger peak in $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insenstive to $B_z$ for approximately 12 hours prior to onset.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{figures/alldata-GOES6-1983-1991.eps}
\caption{Overview of data used in this article. The top four panels show parameters from \cite{Kondrashov2014ReconstructionOfGaps} and the bottom panel contains $\rho_{eq}$ based on GOES 6 measurements from \cite{Denton} after interpolation and averaging described in the text. Dashed horizontal lines in the $D_{st}$ and $\rho_{eq}$ panels indicate sample event cutoff thresholds of $D_{st} = -50$~nT and $\rho_{eq} = 20$~amu/cm$^3$ considered in Section~3.}
\label{AllData}
\end{figure}

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{figures/ccplot-GOES6.eps}
\caption{Top: 27-day non-overlapping medians of $F_{10.7}$ and $log(\rho_{eq})$ from GOES 6. Bottom: Correlation between $log(\rho_{eq})$ and $F_{10.7}$ using medians in non-overlapping hour, day, and 27-day windows.}
\label{ccplot}
\end{figure}
\clearpage

\begin{figure}[h]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dst-50-tak-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dst-day-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using daily medians. Top: Events in the interval 1989-1991; compare to \cite{Takahashi2010} Figure~11. Bottom: Events in the interval 1983-1991.}
\label{DailyAveragedDstEvents}
\end{figure}

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dd12-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dst-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events with the constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.  Bottom: Same as Top except without constraint.}
\label{HourlyAveragedDstEvents}
\end{figure}

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-mass-gt20-GOES6.eps}
\caption{$\rho_{eq} >$ 20~amu/cm$^3$ events from GOES 6 using hourly medians.}
\label{HourlyAveragedRhoEvents}
\end{figure}

>>>>

The top panel of Figure~\ref{fig:ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{fig:ccplot} along with that for 1--day and 1--hour medians.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used combined measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites is 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29, respectively.

@@

The stack plot in the upper panel of Figure \ref{fig:DailyAveragedDstEvents} shows the epoch averages for the 75 events from 1989-1991.  The minimum $D_{st}$ median is $-75$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $4.5$ amu/cm$^3$ instead of the increase of ${\sim} 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  Error lines for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values. For the non-$\rho_{eq}$\ parameters, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of $\rho_{eq}$\ values as all available measurements were used instead of restricting to only values where a $\rho_{eq}$ value also existed.

In the lower panel of Figure \ref{fig:DailyAveragedDstEvents}, it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6 (1983-1991) - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.  This trend is also seen for GOES 7 (1987-1991), is much less significant for GOES 5 (1983-1987), and much more significant for GOES 2 (1981-1983), correlating with the average value of $F_{10.7}$ during each respective coverage period.

To determine if the magnitude of the density medians in the top panel of Figure~\ref{fig:DailyAveragedDstEvents} on epoch days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap sample of the medians for each epoch day was created by sampling the values used to compute the median on each epoch day with replacement.  The observed difference between the medians is shown in Table~\ref{fig:BootstrapDifferenceTable} along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.

@@

A similar hypothesis test run for the bottom panel of Figure~\ref{fig:DailyAveragedDstEvents} yields a maximum confidence level of $\sim$85\%.

@@

\begin{table}
\small
\centering
\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of means of $\rho_{eq}$ shown in the top panel of Figure~\ref{fig:DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

We have also considered the hourly averages from the events shown in the bottom panel of Figure~\ref{fig:DailyAveragedDstEvents}, which shown in the top panel of Figure~\ref{fig:HourlyAveragedDstEvents}.  In the time period of elevated mass density within 6 hours of the start of the event, there are only $\sim$4 samples per hourly interval and the peak value of $\sim$70 amu/cm$^3$ is associated with a single observation.

To increase the number of events, we have removed the constraint that the events must occur in the MLT range from 06:00-12:00 and the constraint that $D_{st}$ remained below the threshold value for 12 hours or longer.  The result is shown in the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents}. The two spikes (one upwards after onset, one downwards before onset) were tested for significant deviation from two baselines, one 24-12 hours before onset, one 16-48 hours after onset. The probability of the left spike coming from the same distribution as the left baseline is 1.06\%, and 8.2\% from the right baseline, whereas the probability of the right spike coming from the left baseline is 1.89\% and 0.17\% from the right baseline.  This significant result is also maintained for GOES 2, 5, and 7. This indicates that there is a statistically significant peak in $\rho_{eq}$ within 6 hours of onset when hourly averages are considered.  This result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

To determine if there is an IMF $B_{z}$ dependence, the epoch curves in the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents} for $\rho_{eq}$ and $B_{z}$ were separated into two parts depending whether $B_z$ was above or below its average four hours before and including the onset hour and four hours after and including the hour of onset; the result is shown in Figure~\ref{fig:BzBinnedDstEvents}. Any statistically significant difference at a 95\% confidence level between the two parts is indicated by a green dot on the horizontal axis using the same bootstrap method described previously.  (At this confidence level, false positives are expected to randomly occur 5\% of the time, or 3.6 times for 72 values.)  These two plots indicate that there is not a statistically significant dependence on IMF $B_{z}$ around the time of the onset of a $D_{st}$ event.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho_{eq}$.  Figure~\ref{fig:HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\rho_{eq}$ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, missing values were replaced with linearly interpolated values.  These events are associated with positive $B_z$ and low values of $D_{st}$.  Approximately 12 hours after onset, $\rho_{eq}$ slowly decays over 40 hours to near pre--onset levels.

Figure~\ref{fig:HourlyAveragedRhoEvents} can be compared with Figure~9 of \cite{Denton2016}, which shows daily averages of $\rho_{eq}$ from this dataset after the onset of quiet $K_p$ intervals.  The growth shown here is approximately linear and lasts $\sim$ 12--18 hours whereas the 10 quiet $K_p$ events considered by \cite{Denton2016} showed growth that lasted for at least 48~hours.

Figure~\ref{fig:BzBinned} shows the the events of Figure~\ref{fig:HourlyAveragedRhoEvents} when separated by $B_z$ in the same manner as Figure~\ref{fig:BzBinnedDstEvents}.  The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to a larger peak in $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insenstive to $B_z$ for approximately 12 hours prior to onset.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{figures/alldata-GOES6-1983-1991.eps}
\caption{Overview of data used in this article. The top four panels show parameters from \cite{Kondrashov2014ReconstructionOfGaps} and the bottom panel contains $\rho_{eq}$ based on GOES 6 measurements from \cite{Denton} after interpolation and averaging described in the text. Dashed horizontal lines in the $D_{st}$ and $\rho_{eq}$ panels indicate sample event cutoff thresholds of $D_{st} = -50$~nT and $\rho_{eq} = 20$~amu/cm$^3$ considered in Section~3.}
\label{fig:AllData}
\end{figure}

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{figures/ccplot-GOES6.eps}
\caption{Top: 27-day non-overlapping medians of $F_{10.7}$ and $log(\rho_{eq})$ from GOES 6. Bottom: Correlation between $log(\rho_{eq})$ and $F_{10.7}$ using medians in non-overlapping hour, day, and 27-day windows.}
\label{fig:ccplot}
\end{figure}
\clearpage

\begin{figure}[h]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dst-50-tak-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dst-day-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using daily medians. Top: Events in the interval 1989-1991; compare to \cite{Takahashi2010} Figure~11. Bottom: Events in the interval 1983-1991.}
\label{fig:DailyAveragedDstEvents}
\end{figure}

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dd12-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dst-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events with the constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.  Bottom: Same as Top except without constraint.}
\label{fig:HourlyAveragedDstEvents}
\end{figure}

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-mass-gt20-GOES6.eps}
\caption{$\rho_{eq} >$ 20~amu/cm$^3$ events from GOES 6 using hourly medians.}
\label{fig:HourlyAveragedRhoEvents}
\end{figure}
